Model details and results

### 2023-09-20 notebook/model-details.qmd
Add mathy notation + prior plots

```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(targets)
library(scales)
library(glue)
library(gt)
library(gtExtras)
library(modelsummary)

@@

theme_set(theme_ongo())
```

>>>>

```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(targets)
library(scales)
library(glue)
library(gt)
library(gtExtras)
library(modelsummary)
library(ggtext)
library(brms)
library(tidybayes)
library(patchwork)

@@

theme_set(theme_ongo())
set_annotation_fonts()
```

# Model definition

$$
\begin{align}
&\ \textbf{Registered provinces for INGO } i \\
\text{Count of provinces}\ \sim&\ \operatorname{Ordered\,Beta}(\mu_{i_j}, \phi_y, k_{1_y}, k_{2_y}) \\[8pt]
&\ \textbf{Model of outcome average} \\
% Put the huge equation in a nested \begin{aligned}[t] environment so that 
% \mathrlap{} can go around it so that the annotations in the priors can be 
% aligned closer to the math
\mu_i =&\ 
  \mathrlap{\begin{aligned}[t]
  & \beta_0 + \beta_1\ \text{Issue[Arts and culture]} + \beta_2\ \text{Issue[Education]}\ +\\
  & \beta_3\ \text{Issue[Industry association]} + \beta_4\ \text{Issue[Economy and trade]}\ + \\
  & \beta_5\ \text{Issue[Charity and humanitarian]} + \beta_6\ \text{Issue[General]}\ + \\
  & \beta_7\ \text{Issue[Health]} + \beta_8\ \text{Issue[Environment]}\ + \\
  & \beta_9\ \text{Issue[Science and technology]} + \beta_{10}\ \text{Local connections}\ + \\
  & \beta_{11}\ \text{Time since Januay 2017} + \beta_{12}\ \text{Year registered}
  \end{aligned}}\\[8pt]
&\ \textbf{Priors} \\
\beta_0\ \sim&\ \operatorname{Student\,t}(\nu = 3, \mu = 0, \sigma = 2.5) && \text{Intercept} \\
\beta_{1..12}\ \sim&\ \mathcal{N}(0, 5) && \text{Coefficients} \\
\phi_y\ \sim&\ \operatorname{Exponential}(1 / 100) && \text{Variability in province count} \\
k_{1_y}, k_{2_y}\ \sim&\ \operatorname{Induced\,Dirichlet}(1, 1, 1), && \text{0–continuous and continuous–1 cutpoints} \\
&\ \quad \text{or } \bigl[P(\alpha_1), P(\alpha_1 + \alpha_2)\bigr] && \quad\text{(boundaries between 3 Dirichlet columns)}
\end{align}
$$

# Model priors

These are all pretty standard. I'd like to use a Student t for the coefficients in addition to the intercept, but the prior distributional families are baked into `ordbetareg()` and the only way to change them is to create a custom {brms} family based on the code in `ordbetareg()` and that's a lot of hassle. So we'll live with a normal distribution for the coefficients.

The only new-to-me distribution here is the Dirichlet prior for the cutpoints, so I wrote a whole blog post about it here. This is a little confusing, though. `ordbetareg()` requires a three-element Dirichlet distribution that contains three columns of probabilities, one per submodel: the only-zero part, the between-zero-and-one part, and the only-one part. However, it doesn't use these three probabilities directly. Instead, it uses an [*induced* Dirichlet distribution](https://betanalpha.github.io/assets/case_studies/ordinal_regression.html) [@Kubinec:2022, p. 524], meaning that it works with the *boundaries between the categories* and not the Dirichlet columns themselves ([see here for more about that, plus some neat visualizations of the cutpoints](https://www.andrewheiss.com/blog/2023/09/18/understanding-dirichlet-beta-intuition/#bonus-later-addition-boundaries-between-categories)).

$\operatorname{Dirichlet}(1, 1, 1)$ represents a uniform distribution of three related columns in a matrix, each with a probability of 33%, since the sum of the three probabilities must be 100%.

$$
\begin{align}
\textbf{E}(\alpha_1) &= \frac{\alpha_1}{\sum{\alpha}} = \frac{1}{1 + 1 + 1} = \frac{1}{3} = 0.333 \\[8pt]
\textbf{E}(\alpha_2) &= \frac{\alpha_2}{\sum{\alpha}} = \frac{1}{1 + 1 + 1} = \frac{1}{3} = 0.333 \\[8pt]
\textbf{E}(\alpha_3) &= \frac{\alpha_3}{\sum{\alpha}} = \frac{1}{1 + 1 + 1} = \frac{1}{3} = 0.333
\end{align}
$$

We can find the distributions of the boundaries between these columns too—$k_1$, or the boundary between the $\alpha_1$ and $\alpha_2$ columns, is immediately after $\alpha_1$, or at 0.33; $k_2$, or the boundary between $\alpha_2$ and $\alpha_3$, is immediately after $\alpha_2$, which is at ($\alpha_1 + \alpha_2$), or 0.66. Because Dirichlet is just a fancy multivariate version of a Beta distribution, each of the $\alpha$ columns by itself is a Beta distribution, and by extension, the $k$ boundaries between the $\alpha$ columns are also Beta distributions:

$$
\begin{align}
k_1 &= \textbf{E}(\alpha_1) = \frac{1}{3} & k_2 &= (\textbf{E}(\alpha_2) + \textbf{E}(\alpha_2)) = \frac{2}{3}\\
&= \operatorname{Beta}(1, 2) & &= \operatorname{Beta}(2, 1)
\end{align}
$$

So here's what all our priors look like:

```{r plot-priors}
prior_summary(m_full_ordbeta) |>
  bind_rows(
    tibble(
      prior = c("beta(1, 2)", "beta(2, 1)"),
      class = c("k1", "k2")
    )
  ) |>
  parse_dist() |>
  filter(
    prior != "",
    !str_detect(prior, "^induced")
  ) |>
  mutate(class_nice = case_match(
    class,
    "Intercept" ~ "β<sub>0</sub>",
    "b" ~ "β<sub>1–12</sub>",
    "phi" ~ "φ",
    "k1" ~ "k<sub>1</sub>",
    "k2" ~ "k<sub>2</sub>"
  )) |>
  mutate(prior_nice = str_to_sentence(str_replace(prior, "_", " "))) |>
  mutate(class = factor(class, levels = c("Intercept", "b", "phi"))) |>
  arrange(class) |>
  mutate(nice_title = glue("**{class_nice}**: {prior_nice}")) %>%
  mutate(nice_title = fct_inorder(nice_title)) %>%
  ggplot(aes(y = 0, dist = .dist, args = .args, fill = prior)) +
  stat_slab(normalize = "panels") +
  scale_x_continuous(labels = label_comma(style_negative = "minus")) +
  scale_fill_manual(values = clrs[c(6, 8, 5, 3, 1)]) +
  facet_wrap(vars(nice_title), scales = "free_x", nrow = 2) +
  guides(fill = "none") +
  labs(
    x = NULL, y = NULL,
    caption = "k<sub>1</sub> and k<sub>2</sub> are the cutpoints or category boundaries from Dirichlet(1, 1, 1)"
  ) +
  theme_ongo(prior = TRUE) +
  theme(
    strip.text = element_markdown(), 
    plot.caption = element_markdown(hjust = 0)
  )
```
